Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-10-22 paper/type-system.tex
Text on static semantics.

Like methods, anonymous functions can also be region-polymorphic. 
% A lambda expression defines a region-polymorphic
% multi-argument function closure parameterized over function's
% allocation context parameter. 
The angle braces in the lambda expression ($\inang{\rho^a\rhobar \,|\,
\phi}$) serve the same purpose as they do in a method definition - to
capture region parameters along with their constraints. Region
parameters also appear in the arrow type of the lambda expression 
($\inang{\rhoalloc\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn} 
\tau$) at the prenex position, similar to ML type schemes. However,
unlike ML, we don't distinguish between region types and region type
schemes; any of the $\tau$'s in the arrow type can themselves be
region-parametric arrow types. In this respect, our region type system
is more like System F's type system, which admits higher-rank
parametric polymorphism. Like System F, \FB provides a region
instantiation expression ($e\inang{\ralloc\rbar}$), and a region
generalization expression ($\Lambdaexp{\rhoalloc\rhobar \,|\,
\phi}{e}$) to instantiate and generalize region variables.

@@

Static semantics of \fbname is defined by the rules to establish
well-formedness of region types and region constraints, check
subtyping between a pair of region types, and type-check expressions
against region types. Fig.~\ref{fig:fb-staticsem} contains an
illustrative subset of such rules\footnote{The full formal development
can be found in our technical report~\cite{techrep}}, along with some
auxiliary definitions. The judgments defined by these rules make use
of a context ($\A$), which is a tuple of:
\begin{itemize}
\item A set ($\rhoenv \in 2^\rgn$) of static identifiers of regions
that are estimated to be live,
\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
variables to their bounds\footnote{A bound of a type variable in
FGJ~\cite{fgj} is the class that type variable was declared to extend.
The map from type variables to their bounds is denoted as $\Delta$
in FGJ formalism}.
\item A constraint formula ($\phicx$) that captures the outlives
constraints on regions in $\rhoenv$.
\end{itemize}

>>>>

Like methods, anonymous functions can also be region-polymorphic. 
% A lambda expression defines a region-polymorphic
% multi-argument function closure parameterized over function's
% allocation context parameter. 
The angle braces in the lambda expression ($\inang{\rho^a\rhobar \,|\,
\phi}$) serve the same purpose as they do in a method definition - to
capture region parameters along with their constraints. Region
parameters also appear in the arrow type of the lambda expression 
($\inang{\rhoalloc\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn}
\tau$) at the prenex position, similar to ML type schemes. However,
unlike ML, we don't distinguish between types and type schemes; any of
the $\tau$'s in the arrow type can themselves be region-parametric
arrow types. In this respect, our region type system is more like
System F's type system, which admits higher-rank parametric
polymorphism. Like System F, \FB provides a region instantiation
expression ($e\inang{\ralloc\rbar}$), and a region generalization
expression ($\Lambdaexp{\rhoalloc\rhobar \,|\, \phi}{e}$) to
instantiate and generalize region variables.

@@

Static semantics of \fbname is defined by the rules to establish
well-formedness of region types and region constraints, check
subtyping between a pair of region types, and type-check expressions
against region types. Fig.~\ref{fig:fb-staticsem} contains an
illustrative subset of such rules\footnote{The full formal development
can be found in our technical report~\cite{techrep}}, along with some
auxiliary definitions. The judgments defined by these rules make use
of a context ($\A$), which is a tuple of:
\begin{itemize}
\item A set ($\rhoenv \in 2^\rgn$) of static identifiers of regions
that are estimated to be live,
\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
variables to their bounds\footnote{A bound of a type variable
($\tyvar$) in FGJ~\cite{fgj} is the class ($\fgjN$) that type variable
was declared to extend.  The map $\tyvar \mapsto \fgjN$ is denoted
$\Delta$ in FGJ formalism}, and
\item A constraint formula ($\phicx$) that captures the outlives
constraints on regions in $\rhoenv$.
\end{itemize}
The context for the expression typing judgment also includes:
\begin{itemize}
\item A type environment ($\env \in x \mapsto \tau$) that contains the
type bindings for variables in scope, and 
\item The static identifier ($\rgn^a$) of the allocation context for
the expression is being type-checked.  
\end{itemize}

\noindent The well-formedness judgment on region types
($\tywf{\A}{\tau}$) makes use of the well-formedness and subtyping
judgments on core types \footnote{We use a double-piped turnstile
($\Vdash$) for judgments in FGJ~\cite{fgj}, and a simple turnstile
($\vdash$) for those in \FB.}.  An existentially quantified type of a
region handler is well-formed if, effectively, $\RgnZ\inang{T}$ is a
well-formed FGJ type. A region type
($B\inang{\ralloc\rbar}\inang{\tbar}$) of an object of class $B$ is
well-formed if and only if (a). its region-erased variant
($B\inang{\tbar}$) is well-formed, and (b). its regions arguments
($\ralloc\rbar$) are live, and satisfy the outlives constraints
($\phi$) required by the class $B$. Well-formedness of region
constraints is straightforward.

The subtype relation between region types of objects
($B\inang{\ralloc\rbar}\inang{\tbar}$) follows straightforwardly from
the subclass relation between their class types. Conspicuous by its
absence among the rules that define subtype relation is a rule that
lifts outlives relation between regions to subtype relation between
types of objects allocated in those regions: 
\begin{smathpar}
\begin{array}{c}
\renewcommand*{\arraystretch}{1.1}
\RULE
  {
    \isvalid{\A.\phicx}{\rgn_1 \outlives \rgn_2}
  }
  {
    \subtyp{\A}{T@{\rgn_1}}{T@{\rgn_2}}
  }
\end{array}
\end{smathpar}
Such a rule is unsound for it allows an object in a region ($\rgn_1$)
to be treated as an object in a shorter living region ($\rgn_2
\underlives \rgn_1$), thus allowing it to refer to shorter living
objects in $\rgn_2$. The type system should prevent such unsafe
references by restricting subtype relation to only those region types
that are invariant in their allocation region. For instance, assuming
that class $A$ is a subclass of $\ObjZ$ (as per FGJ), the subtype
relation $A@\rgn <: \ObjZ@\rgn$ should be allowed, but not the
relation $A@\rgn <: \ObjZ@\rgn'$, for any $\rgn' \neq \rgn$. \FB's
type system enforces this not by restricting the subtype rule, but by
constraining the subclass relation between region-annotated classes.
In particular, it lets a class $A\inang{\rhoalloc}$ to be declared to
extend $\ObjZ\inang{\rhoalloc}$, but never to extend
$\ObjZ\inang{\rhoalloc_1}$, where $\rhoalloc_1 \neq \rhoalloc$. The
well-formedness rule on class definitions (not shown) captures this
restriction.

As expected, subtyping between arrow types is contravariant in
argument types and covariant in return types. It is also contravariant
in the region constraints that the region parameters have to satisfy.
For example, the type $\inang{\rhoalloc\rho} A@\rho \xrightarrow{\rgn}
\unitZ$ is a subtype of $\inang{\rhoalloc\rho \,|\, \rho =  \rhoalloc}
A@\rho \xrightarrow{\rgn} \unitZ$, for the former's precondition is
weaker than the later's. Note that, like other types, the subtype
relation between function types is invariant with respect to the
allocation region ($\rgn$) of their closures. This subtype rule
assumes that both arrow types have same set of region parameters.
However, this is not restrictive as (a). it is possible to add more
region parameters to a function type and declare them to be equal to
existing ones (as the arrow type above does), and (b). it is possible
to rename region parameters by composing instantiation and
generalization (big-lambda) expressions together.

The type rule for \C{letregion} expression requires that the static
identifier introduced by the expression be unique under the current
context (i.e., $\rgn \notin \rhoenv$). This condition is needed in
order to prevent the new region from incorrectly assuming any existing
outlives relationships on an eponymus region. Provided this is
satisfied, the expression ($e$) under \C{letregion} is then
type-checked against $\unitZ$ assuming that the new region is live
($\rgn \in \rhoset$) and that it underlives all existing live regions
($\rhoset \outlives \rgn$).

The \C{new} $\RgnZ\inang{T}$ expression creates a new transferable
region, whose root object is of (core) type $T$, and constructs a new
$\RgnZ$ object to serve as a handler to the region. The corresponding
type rule requires that the type $\RgnZ\inang{T}$ be well-formed as
per FGJ semantics. Furthermore, the constructor is expected to be
passed a thunk that returns an object of type $T$ contained in its
allocation context. For the reasons described earlier, the allocation
region of a new $\RgnZ$ object is existentially quantified.

Like the rule for \C{letregion}, the type rule for the \C{open}
expression requires that the static identifier ($\rgn$) introduced to
refer to the transferable region being opened be unique. However,
unlike the \C{letregion} rule, the rule for \C{open} does not
introduce any outlives relationship between the newly opened region
and any pre-existing region while checking the type of the expression
($e$) under \C{open}. This effectively preempts the possibility of any
references from objects inside the transferable region to those
outside. Environment ($\env$) is extended with binding for the type of
root object while type-checking $e$. The type of region handler ($x$)
is also updated to reflect the elimination of existentially quantified
region variable ($\rho$) in favor of concrete identifier ($\rgn$).

The type rule for lamdba expression requires that its region
parameters ($\rhoalloc\rhobar$) be unique under the current context,
and that the constraints ($\phi$) over region parameters be
well-formed. The lambda-bound expression ($e$) is checked under an
extended type environment containing bindings for function's
arguments, assuming that region parameters are live, and that declared
constraints over region parameters hold. Note that the closure is
always allocated in the current allocation context ($\rgn^a$). This
prevents the closure from escaping the context in which it is created
(cf. the following paragraph), thus trivially ensuring the safety of
any dereferences inside the closure.

The type rule for assignment expression ($e_1 := e_2$) ensures that a
(local or instance) variable is only assigned objects of either the
same type or a subtype. Recall that subtyping between region types is
invariant with respect to the allocation regions of their objects. It
follows that if a variable is declared to refer to objects in a region
$\rgn$, then it is only ever assigned objects that allocated in
$\rgn$. This prevents, for example, a temporary object allocated
inside a stack region within a method from escaping the region's scope
via an assignment to an instance variable.

\subsection{Operational Semantics and Type Safety}

Operational semantics of \fbname defines an evaluation relation
($\redsto$) over its expression language. In particular, the
evaluation gets stuck if an expression attempts to dereference a
pointer into a region that has either been deallocated, freed or
transferred. On the other hand, if expression attempts to commit an
operation on a transferable region handler (a $\RgnZ$ object) that is
not sanctioned by the state transition discipline in
Fig.~\ref{fig:region-fsm}, then it is reduced to a value called
\invalidexn. The accompanying technical report~\cite{techrep} contains
full operational semantics of \fbname.
